Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-07-09 paper.tex
updated to re-submitted version

\begin{abstract}
{\bf
The vertical shear instability (VSI) offers a potential hydrodynamic mechanism to drive angular momentum transport in cold protoplanetary disks (PPDs). The VSI is driven by a weak vertical gradient in the disk's orbital motion, but must overcome vertical buoyancy, a strongly
  stabilizing influence in cold disks, where
  heating is dominated by external irradiation.  Rapid radiative cooling      reduces the effective buoyancy
  and allows the VSI to operate.  We quantify the
  cooling timescale, $t_c$, needed for growth of the VSI. We perform a linear analysis of the VSI with cooling in
  vertically global, radially local disk models. We find the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial
  power-law temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer $t_c$, the VSI
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to
  viscous or turbulent decay.  We apply our results to PPD models
  where $t_c$ is determined by the opacity of dust grains.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI.
}

>>>>

\begin{abstract}
{\bf
The vertical shear instability (VSI) offers a potential hydrodynamic mechanism for angular momentum transport in protoplanetary disks (PPDs). The VSI is driven by a weak vertical gradient in the disk's orbital motion, but must overcome vertical buoyancy, a strongly stabilizing influence in cold disks, where heating is dominated by external irradiation.  Rapid radiative cooling reduces the effective buoyancy and allows the VSI to operate.  We quantify the cooling timescale $t_c$ needed for efficient VSI growth, through a linear analysis of the VSI with cooling in vertically global, radially local disk models. We find the VSI is most vigorous for rapid cooling with $t_c<\OmK^{-1}h|q|/(\gamma -1)$ in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect-ratio, $h\ll1$; the radial power-law temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer $t_c$, the VSI  is much less effective because growth slows and shifts to smaller length scales, which are more prone to viscous or turbulent decay.  We apply our results to PPD models where $t_c$ is determined by the opacity of dust grains.  We find that the VSI is most effective at intermediate radii, from $\sim5$AU to $\sim50$AU with a characteristic growth time of $\sim30$ local orbital periods. Growth is suppressed by long cooling times both in the opaque inner disk and the optically thin outer disk.  Reducing the dust opacity by a factor of 10 increases cooling times enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.
}
